---
title: "Data Analysis Assignment #2 (75 points total)"
author: "Quiballo, Kagen"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

### Instructions

R markdown is a plain-text file format for integrating text and R code, and creating transparent, reproducible and interactive reports. An R markdown file (.Rmd) contains metadata, markdown and R code "chunks", and can be "knit" into numerous output types. Answer the test questions by adding R code to the fenced code areas below each item. There are questions that require a written answer that also need to be answered. Enter your comments in the space provided as shown below:

***Answer: (Enter your answer here.)*** 

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded and a feedback report returned with comments.  Points assigned to each item appear in the template.

**Before proceeding, look to the top of the .Rmd for the (YAML) metadata block, where the *title*, *author* and *output* are given. Please change *author* to include your name, with the format 'lastName, firstName.'**

If you encounter issues with knitting the .html, please send an email via Canvas to your TA.

Each code chunk is delineated by six (6) backticks; three (3) at the start and three (3) at the end. After the opening ticks, arguments are passed to the code chunk and in curly brackets. **Please do not add or remove backticks, or modify the arguments or values inside the curly brackets**. An example code chunk is included here: 

```{r exampleCodeChunk, eval = FALSE, echo = TRUE}
# Comments are included in each code chunk, simply as prompts

#...R code placed here

#...R code placed here

```

R code only needs to be added inside the code chunks for each assignment item. However, there are questions that follow many assignment items. Enter your answers in the space provided. An example showing how to use the template and respond to a question follows.

-----

**Example Problem with Solution:**

Use *rbinom()* to generate two random samples of size 10,000 from the binomial distribution. For the first sample, use p = 0.45 and n = 10. For the second sample, use p = 0.55 and n = 10. Convert the sample frequencies to sample proportions and compute the mean number of successes for each sample. Present these statistics.

```{r Example, eval = TRUE, echo = TRUE}

set.seed(123)
sample.one <- table(rbinom(10000, 10, 0.45)) / 10000
sample.two <- table(rbinom(10000, 10, 0.55)) / 10000

successes <- seq(0, 10)

round(sum(sample.one*successes), digits = 1) # [1] 4.5
round(sum(sample.two*successes), digits = 1) # [1] 5.5
```

**Question: How do the simulated expectations compare to calculated binomial expectations?**

***Answer:  The calculated binomial expectations are 10(0.45) = 4.5 and 10(0.55) = 5.5.  After rounding the simulated results, the same values are obtained.***

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("/Users/kagenquiballo/Desktop/Grad school/MSDS 401/Week 9/da assignment 2/mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1: (5 points) ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}
par(mfrow=c(1,2))
hist(mydata$RATIO)
qqnorm(mydata$RATIO)
qqline(mydata$RATIO)
par(mfrow=c(1,1))
rockchalk::skewness(mydata$RATIO)
rockchalk::kurtosis(mydata$RATIO) + 3
```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}

mydata$L_RATIO <- log10(mydata$RATIO)

par(mfrow=c(1,2))
hist(mydata$L_RATIO)
qqnorm(mydata$L_RATIO)
qqline(mydata$L_RATIO)
par(mfrow=c(1,1))
rockchalk::skewness(mydata$L_RATIO)
rockchalk::kurtosis(mydata$L_RATIO) + 3
boxplot(mydata$L_RATIO ~ mydata$CLASS)
title("boxplot of L_RATIO by CLASS")

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
bartlett.test(x=mydata$RATIO, g=mydata$CLASS)
bartlett.test(x=mydata$L_RATIO, g=mydata$CLASS)
```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: RATIO has a skewness=0.7147056, kurtosis=4.667298, and p-value=0.0002531 for test of homogeneity of variance. Because the p-value is < .05, we conclude the variance of RATIO are not homogenous across age classes. Comparitively, L_RATIO has a skewness=-0.09391548, kurtosis=3.535431, and p-value=0.5267 for test of homogeneity of variance. Because the p-value is > .05, we fail to reject the hypothesis of homogenous variances of L_RATIO by age class. Because L_RATIO has a skewness closer to 0 and a kurtosis closer to 3, it conforms better to a normal distribution. ***


***#### Section 2 (10 points) ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
no_interaction <- aov(L_RATIO~CLASS+SEX, data=mydata)
interaction <- aov(L_RATIO~CLASS+SEX+CLASS:SEX, data=mydata)

summary(no_interaction)
summary(interaction)
```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: For both models, CLASS and SEX individually are stasticially significant and will play a role in predicting L_RATIO. From the interaction model, the interaction term CLASS:SEX is not significant. This means that the joint effect of CLASS and SEX is not statistically higher than the sum of the individual effects when predicting L_RATIO. In other words, there is no significant interaction between SEX and CLASS that affects L_RATIO.***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(no_interaction)
```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: For age classes, the Tukey HSD compares older minus younger. The difference for all age class comparisons (with the exception of A2-A1) is negative and does not include 0 in the 95% confidence interval. This indicates that the lower age classes (younger) typically have higher L_RATIOS than higher age classes (older). For sexes, M-F does contain 0 in the 95% confidence interval which indicates no significant difference. They can indeed be combined into a single category named "adults."***


***####  Section 3: (10 points) ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:
mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)
```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}
par(mfrow=c(1,2))
hist(mydata$VOLUME[mydata$TYPE=="I"])
#median(mydata$VOLUME[mydata$TYPE=="I"]) #148.4515
hist(mydata$VOLUME[mydata$TYPE=="ADULT"])
#median(mydata$VOLUME[mydata$TYPE=="ADULT"]) #388.7189
par(mfrow=c(1,1))
```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The infants have a highly skewed distribution (right) for VOLUME, whereas the VOLUME of adults are approximately normal. The median VOLUME of infant is 148 whereas the median VOLUME of adults is 388 making differentiation of the distributions of VOLUME by TYPE easier.***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}
par(mfrow=c(2,2))

mydata$L_SHUCK <- log10(mydata$SHUCK)
mydata$L_VOLUME <- log10(mydata$VOLUME)

plot(SHUCK~VOLUME, data=mydata, col=CLASS)
legend("topleft", legend=levels(mydata$CLASS), pch=16, col=unique(mydata$CLASS))
title("SHUCK vs VOLUME (by CLASS)")

plot(L_SHUCK~L_VOLUME, data=mydata, col=CLASS)
legend("topleft", legend=levels(mydata$CLASS), pch=16, col=unique(mydata$CLASS))
title("L_SHUCK vs L_VOLUME (by CLASS)")

plot(SHUCK~VOLUME, data=mydata, col=TYPE)
legend("topleft", legend=levels(mydata$TYPE), pch=16, col=unique(mydata$TYPE))
title("SHUCK vs VOLUME (by TYPE)")

plot(L_SHUCK~L_VOLUME, data=mydata, col=TYPE)
legend("topleft", legend=levels(mydata$TYPE), pch=16, col=unique(mydata$TYPE))
title("L_SHUCK vs L_VOLUME (by TYPE)")

par(mfrow=c(1,1))
```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: There is a positive linear relationship between L_SHUCK and L_VOLUME. For CLASS, lower age classes tend to have lower values of L_SHUCK and L_VOLUME whereas higher age classes have higher values. On the contrary, for TYPE, infants tend to have higher values of L_SHUCK and L_VOLUME whereas adults have lower values. In regards to variability, L_SHUCK and L_VOLUME now have equal variance after a logarithmic transformation (versus the previous increasing variance). Equal variance of errors for a linear regression analysis implies efficient and unbiased estimators.***


***####   Section 4: (5 points) ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}
model4 <- lm(formula = L_SHUCK ~ L_VOLUME + CLASS + TYPE, data=mydata)
summary(lm(formula = L_SHUCK ~ L_VOLUME + CLASS + TYPE, data=mydata))

```

**Essay Question:  Interpret the trend in CLASS level coefficient estimates? (Hint: this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: CLASS coefficients have increasingly negative coefficients. Holding all other predictors constant, a higher CLASS will result in a lower L_SHUCK value. However, the scatterplot from above suggested that higher age classes had higher values of L_SHUCK. ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: With the TYPEI coefficient estimate merely being -0.04857, this small change in the value of L_SHUCK pales in comparison to CLASS and L_VOLUME which are about 10x greater in magnitude. We can conclude that TYPE is not an imporant predictor in this regression.***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: (5 points) ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}
par(mfrow=c(1,2))
hist(model4$residuals)
qqnorm(model4$residuals)
qqline(model4$residuals)
par(mfrow=c(1,1))
rockchalk::skewness(model4$residuals)
rockchalk::kurtosis(model4$residuals) + 3
```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}
par(mfrow=c(2,2))

plot(mydata$L_VOLUME, model4$residuals, col=mydata$CLASS)
legend("topleft", legend=levels(mydata$CLASS), pch=16, col=unique(mydata$CLASS))
title("L_SHUCK vs residuals (by CLASS)")

plot(mydata$L_VOLUME, model4$residuals, col=mydata$TYPE)
legend("topleft", legend=levels(mydata$TYPE), pch=16, col=unique(mydata$TYPE))
title("L_SHUCK vs residuals (by TYPE)")

boxplot(model4$residuals~mydata$CLASS, col=mydata$CLASS)
legend("topleft", legend=levels(mydata$CLASS), pch=16, col=unique(mydata$CLASS))
title("L_SHUCK vs residuals (by CLASS)")

boxplot(model4$residuals~mydata$TYPE, col=mydata$TYPE)
legend("topleft", legend=levels(mydata$TYPE), pch=16, col=unique(mydata$TYPE))
title("L_SHUCK vs residuals (by TYPE)")

par(mfrow=c(1,1))
```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: From the skewness, kurtosis, and qqplot, we can conclude that the variance of errors is equal and normally distributed. The residuals are very similar across different values of CLASS and TYPE which shows the model is a good predictor of L_VOLUME and ultimately VOLUME for harvesting decisions.***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: (5 points) ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383

# Proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /
  total.adults  # [1] 0.2476573

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) / 
  total.infants #0% of infants harvested

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:
(med_inf_vol <- median(mydata$VOLUME[mydata$TYPE == "I"]))  # 133.8214

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) / total.infants #49.8% of infants harvested

# Add code to calculate the proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) / total.adults  # 93.3% of adults harvested

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.

# Add code to determine the median adult volume:
(med_ad_vol <- median(mydata$VOLUME[mydata$TYPE == "ADULT"]))  # 384.5584

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > med_ad_vol) / total.infants # 2.4% of infants harvested

# Add code to calculate the proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_ad_vol) / total.adults  # 49.9% of adults harvested

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}
plot(volume.value,prop.infants,type="o", col="blue")
points(volume.value, prop.adults, col="dark red")
lines(volume.value, prop.adults, col="dark red")
legend("bottomright", legend=c("Infants", "Adults"), pch=16, col=c("blue", "dark red"), cex=2)
abline(v = max_inf_vol)
abline(v = med_inf_vol)
abline(v = med_ad_vol)
abline(h = .5, lty=2)
text(x = 516, y = .2, label = "max infant volume", srt = 90, cex = 1)
text(x = 123, y = .8, label = "median infant volume", srt = 90, cex = 1)
text(x = 374, y = .2, label = "median adult volume", srt = 90, cex = 1)
title("Saved Proportions vs Volume of Abalone")
```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The median infant volume allows half of the infants to be harvested and the majority of adults to be harvested. The median adult volume allows half of the adults to be harvested and barely any of the infants to be harvested. These populations have greatly differing median volumes which will help indicate which ones to harvest. It all depends on what a feasible amount of infants one can leave during harvest to collect the most while still saving enough for future harvests. ***

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

***#### Section 7: (10 points)  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a, warning = FALSE}
difference <- (1 - prop.adults) - (1 - prop.infants)
plot(volume.value, difference, col="light gray")
#max_dif <- loess.smooth(volume.value, difference, degree=100)$x[which.max(loess.smooth(volume.value, difference, degree=100)$y)]
abline(v = max_inf_vol)
abline(v = med_inf_vol)
abline(v = med_ad_vol)
#abline(v = max_dif)
text(x = 516, y = .1, label = "max infant volume", srt = 90, cex = .7)
text(x = 123, y = .1, label = "median infant volume", srt = 90, cex = .7)
text(x = 374, y = .1, label = "median adult volume", srt = 90, cex = .7)
title("(1 - prop.adults) - (1 - prop.infants) vs Volume")
#text(x = 277, y = .1, label = "max difference", srt = 90, cex = 1)

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}
max_dif <- volume.value[which.max(difference) ]

par(mfrow=c(1,2),mar = c(3, 2, 2, 1))
plot(volume.value, difference, col="light gray")
lines(volume.value, smooth.difference, col="dark red", cex=5)
legend("bottomright", legend=c("Raw data", "Smooth data"), pch=16, col=c("light gray", "dark red"), cex=2)
title("Smoothed Difference Plot vs Volume")

plot(volume.value,prop.infants,type="o", col="blue")
points(volume.value, prop.adults, col="dark red")
lines(volume.value, prop.adults, col="dark red", cex=2)
legend("bottomright", legend=c("Infants", "Adults"), pch=16, col=c("blue", "dark red"), cex=2)
abline(v = max_inf_vol)
abline(v = med_inf_vol)
abline(v = med_ad_vol)
abline(v = max_dif)
max_dif <- volume.value[which.max(smooth.difference)]
text(x = 506, y = .2, label = "max infant volume", srt = 90, cex = 1)
text(x = 113, y = .8, label = "median infant volume", srt = 90, cex = 1)
text(x = 364, y = .2, label = "median adult volume", srt = 90, cex = 1)
text(x = max_dif-20, y = .5, label = "maximum difference", srt = 90, cex = 1, col="red")
title("Unharvested Proportions vs Volume")
par(mfrow=c(1,1))

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the volume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
# OR,
sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8: (10 points)  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

sum(mydata[mydata$TYPE == "I", "VOLUME"] > 206.786) / total.infants #29% of infants harvested
sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] > 206.786) / total.adults #83% of infants harvested
```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}
par(mfrow=c(1,1))
plot(volume.value,prop.infants,type="o", col="blue")
points(volume.value, prop.adults, col="dark red")
lines(volume.value, prop.adults, col="dark red")
legend("bottomright", legend=c("Infants", "Adults"), pch=16, col=c("blue", "dark red"), cex=2)
abline(v = max_inf_vol)
abline(v = med_inf_vol)
abline(v = med_ad_vol)
abline(v = max_dif)
abline(v = 206.786)
text(x = 516, y = .2, label = "max infant volume", srt = 90, cex = 1)
text(x = 123, y = .8, label = "median infant volume", srt = 90, cex = 1)
text(x = 374, y = .2, label = "median adult volume", srt = 90, cex = 1)
text(x = max_dif-10, y = .5, label = "maximum difference", srt = 90, cex = 1)
text(x = 206.786-10, y = .5, label = "zero A1 infants", srt = 90, cex = 1, col="red")
title("Unharvested Proportions vs Volume")
par(mfrow=c(1,1))
```


***#### Section 9: (5 points) ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}
max_inf_vol_cut <- volume.value[which.min(abs(volume.value-max_inf_vol))]
med_inf_vol_cut <- volume.value[which.min(abs(volume.value-med_inf_vol))]
med_ad_vol_cut <- volume.value[which.min(abs(volume.value-med_ad_vol))]
no_class_a1_cut <- volume.value[which.min(abs(volume.value-206.786))]

plot((1 - prop.infants),(1 - prop.adults),     
     col= ifelse(volume.value %in% c(max_inf_vol_cut, med_inf_vol_cut, med_ad_vol_cut, max_dif, no_class_a1_cut), "red", "black"),
     cex = ifelse(volume.value %in% c(max_inf_vol_cut, med_inf_vol_cut, med_ad_vol_cut, max_dif, no_class_a1_cut), 4, 1))

text(x = .12, y = .25, label = "max infant volume", cex = 1)
text(x = .57, y = .87, label = "median infant volume", cex = 1)
text(x = .16, y = .5, label = "median adult volume", cex = 1)
text(x = .3, y = .68, label = "maximum difference", cex = 1)
text(x = .35, y = .77, label = "zero A1 infants", cex = 1)
title("ROC Graph")
abline(a=0,b=1,lty=2)
```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
auc((1 - prop.infants),(1 - prop.adults))
```


***#### Section 10: (10 points) ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults),
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	
title <- c("protect all infants","median infant","median adult","max difference","zero A1 infants")
cut_offs <- c(max_inf_vol, med_inf_vol, med_ad_vol, max_dif, 206.786)
true_positive <- c((1-prop.adults)[volume.value==max_inf_vol_cut], 
                   (1-prop.adults)[volume.value==med_inf_vol_cut],
                   (1-prop.adults)[volume.value==med_ad_vol_cut],
                   (1-prop.adults)[volume.value==max_dif],
                   (1-prop.adults)[volume.value==no_class_a1_cut])
false_positive <- c((1-prop.infants)[volume.value==max_inf_vol_cut], 
                   (1-prop.infants)[volume.value==med_inf_vol_cut],
                   (1-prop.infants)[volume.value==med_ad_vol_cut],
                   (1-prop.infants)[volume.value==max_dif],
                   (1-prop.infants)[volume.value==no_class_a1_cut])
harvest <- c(sum(mydata[, "VOLUME"] > max_inf_vol) / (total.infants+total.adults),
             sum(mydata[, "VOLUME"] > med_inf_vol) / (total.infants+total.adults),
             sum(mydata[, "VOLUME"] > med_ad_vol) / (total.infants+total.adults),
             sum(mydata[, "VOLUME"] > max_dif) / (total.infants+total.adults),
             sum(mydata[, "VOLUME"] > 206.786) / (total.infants+total.adults))
data.frame(title,cut_offs,true_positive,false_positive,harvest)
```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: Hypothetically, we want the highest true positive rate (1-prop.adults) and the lowest false positive rate (1-prop.infants). Max infant volume has a poor true positive rate, whereas median infant volume has a high false positive rate -- both rank poorly on the ROC curve. Median adult vloume also has a poor true positive rate, whereas zero A1 infants also has a high false positive rate -- both rank moderately on the ROC curve. The point closest to the top left graph (maximum difference) has an approximately equal amount of trade off between true positives and true negatives. This allows us to harvest adults correctly while also avoiding infants for future harvests.***

**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: I would start off by determining if harvesting at each of the cutoff values is sustainable, i.e harvesting at one cutoff ensures that we will have a population of similar or greater size for the next harvest. From the cutoffs that are sustainable, I would choose the one that achieves the highest proportion of harvested adults while ensuring the population of infants will yield the same (or greater) harvests in the future. Not knowing what proportion of infants needed to sustain a harvest for the following yield poses a limitation to this discussion. However, we can still compare the cutoff points of tradeoffs. I would suggest the max difference because it yields the proportion of adults (74%) harvested that is the greatest amount bigger than the amount of infants harvested (18%). It also yields the best tradeoff between true positives and false positives on the ROC curve. Using this cutoff (volume.value=262.1430) is only a suggestion. Based on past research, I would use this information in addition to 'what is the minimum amount of infants needed to continue sustainable harvesting' and find this point on the ROC curve as a false positives. Another note from this analysis is that correlation does not mean causation. So the results from this study will not necessarily hold true at all times but rather help provide insight on the tradeoffs between correctly harvesting adults vs wrongfully harvesting infants. Lastly, I would suggest that future studies include historical data to compare the population to see if the previously selected cut offs yielded sustainable future harvests. I would also suggest other considerations such as the time of year they were harvested and what climate conditions occurred to get a better sense of what cultivated such a harvest.***